---
title       : UK Student Loan Calculator 
subtitle    : 
author      : Steve Mainprize
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## Introduction

This app lets you see how much a university student in England or Wales is likely to have to repay on their student loan, and for how long. A student typically borrows their tuition fees and some living expenses. Interest is charged at RPI (Retail Price Index) + 3%, and repayments are automatically deducted from the student’s salary once they start earning, at a rate of 9% of everything they earn over £21,000. Any debt not paid off after 30 years is written off.

The app lets you play with these parameters:

* The annual tuition fee charged by the university. Usually it’s £9000 per year.
* What the student wants to borrow per year for living expenses. It’s capped according to household income.
* How many years the course runs for. You can select 3, 4 or 5 years.
* What annual salary the student might get when they graduate and start work.
* An estimate of the rate at which the salary might increase per year.
* Retail Price Index (RPI), an official government measure of inflation. It determines the rate at which the student is charged interest.

--- .class #id 

## Repayments for a student with a starting salary of £25,000

```{r echo=FALSE}

    library(ggplot2)
 
    calcloanmodel<-function(rpi, years, annualtuitionfee, 
                            annualmaintenance, startingsalary, salaryincrease) {
    # Initialise variables

    # Debt brought forward and carried forward
    bf <- 0
    cf <- 0
    
    # Year and month in which course starts. Could be input items, but decided to hard-code
    # them to save space on the page
    startyearofcourse <- 2016
    date <- as.Date(paste(startyearofcourse,"-09-01",sep = ""))
    
    # Monthly interest rate, calculated from the annual RPI figure
    monthlyinterestrate <- (1 + (3 + rpi) / 100) ^ (1 / 12) - 1
    
    # Flags for remembering where we're up to in the status of the loan
    repaymentsstarted <- FALSE
    loanwrittenoff <- FALSE

    # How many months we need in the model    
    monthsneeded <- 12 * (as.numeric(years) + 1 + 30) - 5

    # Create a data frame. This will hold a month-by-month flow of debt based on the parameters    
    loanmodel <-
      data.frame(
        date = as.Date(rep(0,monthsneeded), origin = "1900-01-01"),
        age = numeric(monthsneeded),
        annualsalary = numeric(monthsneeded),
        bf = numeric(monthsneeded),
        receipts = numeric(monthsneeded),
        interestadded = numeric(monthsneeded),
        repayment = numeric(monthsneeded),
        cf = numeric(monthsneeded)
      )
    
    # Initialisation of variables complete
    
    # Start to calculate the flow of the loan model
    
    rowcount <- 0
    while (rowcount < monthsneeded) {
      
      # increment the row count
      rowcount <- rowcount + 1
      
      # record the date that this row represents
      loanmodel$date[rowcount] <- date
      
      # Set the row's "brought forward" to the previous row's "carried forward"
      loanmodel$bf[rowcount] <- cf

      # If it's the start of an academic year, add more debt to the pile      
      if ((format(date,"%m") == "09") &
          (rowcount < as.numeric(years) * 12)) {
        loanmodel$receipts[rowcount] <-
          annualtuitionfee + annualmaintenance
      }
      
      # Calculate the interest added in the current month
      loanmodel$interestadded[rowcount] <-
        (loanmodel$bf[rowcount] + loanmodel$receipts[rowcount]) * monthlyinterestrate
      
      # Repayments start in the April after the student has graduated.
      if ((format(date,"%m") == "04") &
          (rowcount >= as.numeric(years) * 12)) {
        repaymentsstarted <- TRUE
      }
      
      # Work out what the student is earning in the current month
      if (rowcount <= (as.numeric(years) * 12)) {
        # Hasn't started earning yet 
        loanmodel$annualsalary[rowcount] <- 0
      } else if (rowcount == (1 + as.numeric(years) * 12)) {
        # First month of getting paid
        loanmodel$annualsalary[rowcount] <- startingsalary
      } else if (format(date,"%m") == "09") {
        # Assumes that the student gets a pay rise every September, i.e. each anniversary
        # of starting work. Monthly pay packet this month is calculated by adding the assumed
        # annual increase to last month's pay packet.
        loanmodel$annualsalary[rowcount] <-
          round(loanmodel$annualsalary[(rowcount - 1)] * (100 + salaryincrease) / 100, digits =
                  -2)
      } else if (rowcount > (as.numeric(years) * 12)) {
        # The student gets what they got last month
        loanmodel$annualsalary[rowcount] <-
          loanmodel$annualsalary[(rowcount - 1)]
      }
      
      if (repaymentsstarted) {
        # Calculate their debt repayment this month, which is 9% of whatever their salary
        # is over 21000 (per annums converted to per month, obviously)
        if (loanmodel$annualsalary[rowcount]>21000) {
          loanmodel$repayment[rowcount] = 0.09 * (loanmodel$annualsalary[rowcount] -
                                                    21000) / 12
        }
      }
      
      # Work out what debt gets carried forward to next month
      loanmodel$cf[rowcount] <- loanmodel$bf[rowcount] +
        loanmodel$receipts[rowcount] +
        loanmodel$interestadded[rowcount] -
        loanmodel$repayment[rowcount]
      
      # If the debt carried forward to next month would be negative, then the student
      # has repaid the debt. Cap the repayment to the amount outstanding on the loan.
      if (loanmodel$cf[rowcount] < 0) {
        loanmodel$repayment[rowcount] = loanmodel$bf[rowcount] +
          loanmodel$receipts[rowcount] +
          loanmodel$interestadded[rowcount]
        loanmodel$cf[rowcount] = 0
      }
      
      cf <- loanmodel$cf[rowcount]
      
      # add a month to the date
      date <- seq(date, by = paste (1, "months"), length = 2)[2]
      
    }
    
    return(loanmodel)
    
    }
    

  # First Model
  rpi <- 0.9
  years <- 3
  annualtuitionfee <- 9000
  annualmaintenance <- 3821
  startingsalary <- 25000
  salaryincrease <- 4
  lm1 <- calcloanmodel(rpi, years, annualtuitionfee, annualmaintenance, startingsalary, salaryincrease)

  # Draw a plot of the amount of debt outstanding
    plot1 <- ggplot(data=lm1, aes(x=date, y=cf)) +
      ggtitle("Amount of Debt") +
      ylab("Pounds") +
      xlab("Year") +
      theme(legend.position = "none") +
      theme(plot.title = element_text(lineheight=.8, face="bold")) +
      scale_y_continuous(expand = c(0,0)) +
      expand_limits(y = c(0,60000)) + 
      theme(axis.line = element_line(size = 3)) +
      geom_line(aes(group=1), colour="#D55E00")

```

Amount borrowed: £`r format(sum(lm1$receipts),nsmall=2,big.mark=",",small.mark=",") `
Amount repaid: £`r format(sum(lm1$repayment),nsmall=2,big.mark=",",small.mark=",") `

```{r fig.width=7, fig.height=4, echo=FALSE}
  plot1
```

--- .class #id 

## Repayments for a student with a starting salary of £40,000

```{r echo=FALSE}

  # Second Model - exactly the same as the previous model, but with a higher starting salary
  
  startingsalary <- 40000
  lm2 <- calcloanmodel(rpi, years, annualtuitionfee, annualmaintenance, startingsalary, salaryincrease)

  # Draw a plot of the amount of debt outstanding
    plot2 <- ggplot(data=lm2, aes(x=date, y=cf)) +
      ggtitle("Amount of Debt") +
      ylab("Pounds") +
      xlab("Year") +
      theme(legend.position = "none") +
      theme(plot.title = element_text(lineheight=.8, face="bold")) +
      scale_y_continuous(expand = c(0,0)) +
      expand_limits(y = c(0,60000)) + 
      theme(axis.line = element_line(size = 3)) +
      geom_line(aes(group=1), colour="#D55E00")

```

Amount borrowed: £`r format(sum(lm2$receipts),nsmall=2,big.mark=",",small.mark=",") `
Amount repaid: £`r format(sum(lm2$repayment),nsmall=2,big.mark=",",small.mark=",") `

```{r fig.width=7, fig.height=4, echo=FALSE}
  plot2
```

The student with the higher salary repays less in total, because they repay the debt more quickly.

--- .class #id 

## Conclusion

One interesting result that this app illustrates is that the repayment is a more dependent on the student’s salary than how much they borrow. Within certain parameters, a student repays the same amount over thirty years, even given considerable variation in the amount of debt.

Because interest is usually accruing faster that the student is paying it off, the debt will keep growing for many years, and most students will still be paying a large amount of their income towards paying off their student loan into their fifties.

This app will be useful in educating potential students about the commitment that they are about to enter into.




